The shape of kelp: environmental influences on kelp morphology

### 2018-10-19 chapter_2.Rmd
some more house cleaning

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera [@Bolton2010], nevertheless form the basis of one of the most productive ecosystems globally [@Mann1973]. Kelps generally have a dependence on cool, temperate and arctic seawater temperatures [@Santelices2007; @Bolton2010], and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Outside of these latitudes, they are also found in cool, deep water towards the tropics [@Graham2007]: this deeper tropical water, due to its water clarity, also allows for suitable photon flux and a nitrate concentration [@Luning1990; @Dayton1999; @Zimmerman1984], the two other environmental variables that support kelp productivity. Although kelps have a low taxonomic diversity, their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude other turf and subcanopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates [@Mann1973; @Dunton1987; @Duggins1989; @Steneck2002], each depending on a wide suite of ecological services provided by the kelp habitat [@Gaines1987; @Bologna1993; @Levin1994; @Anderson1997].

@@

The biogeographic distribution of kelp is limited by seawater temperature [@Bolton2010], where increasing temperature gradients reduce kelp distribution. Due to this limiting factor, the two main species of kelps in southern African waters, _Ecklonia maxima_ and _Laminaria pallida_, are distributed along a section of the south coast from De Hoop, extending west around the Cape Peninsula, and thriving north into Namibia [@Molloy1990; @Stegenga1997]. This distribution follows a temperate gradient, where sea temperatures increase as one moves south from Namibia, around Cape Point and towards De Hoop. Although the two species occur together for the majority of the coast, their basic morphologies and resource needs vary to a degree. The larger species, _Ecklonia maxima_, is distributed from Lüderitz to Cape Agulhas [@Bolton1985a; @Probyn1985; @Bolton1987; @Bolton2012]. Characterised by a large distal swollen bulb filled with gas, and smooth fronds, this species grows to approximately 10 meters [@Bolton1987]. There was, however, a 17-meter specimen collected in 2015 off Cape Point (Smit, unpubl. data).This species of kelp not only dominate the biomass of the South African nearshore, but plays an important ecological role [@Bustamante1996]. The estimated productivity of _Ecklonia maxima_ within South Africa varies between 350 and 1500g Cm-2yr-1 [@Mann1982]. Across the majority of the coastline, _Laminaria pallida_ remains a subsurface kelp, dominating the kelp biomass at depths greater than 10 meters [@Field1980b; @Bolton1987; @Molloy1996]. This species is distributed from Danger Point, east of the Cape Peninsula, to Rocky Point in northern Namibia, and reaches depths of greater than 20 meters [@Field1980b; @Molloy1990; @Molloy1996; @Stegenga1997]. Towards the north along the west coast, from around Hondeklipbaai, _Laminaria pallida_ replaces _Ecklonia maxima_ as the dominant kelp species [@Velimirov1977; @Stegenga1997] and it also occupies increasingly shallower subtidal regions. The northern populations also exhibit an increase in stipe hollowness, compared to the solid stipe morphs in the species’ southern distributions [@Molloy1996]. This variation in morphology was thought to represent two distinct species, with the northern populations formerly described as _Laminaria schinzii_ Foslie [@Molloy1996]. Genetic work has subsequently shown that the two morphs are in fact the same species [@Rothman2017]. In southern African waters, the primary production of _Laminaria pallida_ is between 120 and 1900g C m2yr1, similar to that of _Ecklonia maxima_ [@Mann1982]. Primary production is not the only pathway.

@@

In Figure 2 the total coastal wave exposure of the Cape Peninsula is given in terms of wave energy (kW per meter wave crest length). These values have been computed using model output from approximately 20 year of simulated nearshore wave conditions. Hindcast global wave parameters were refracted into the nearshore via a spectral phase-averaged numerical wave modelling code SWAN (Simulating, WAves in the Nearshore - 3rd generation). The model is fully spectral in all frequencies and directions (0o -360o) (can solve swell and local wind generated waves simultaneously, propagating in different direction) and was used to model the propagation of short crested swell waves, waves generated by wind, non-linear wave-wave interaction and dissipation and depth induced breaking and refraction. Wave energy dissipation due to bottom friction and whitecapping is also accounted for (Deltares, 2014).

Time series output from 1997 till 2013 of this model was then produced at the 7m and 15m depth contour lines. Each nearshore output point thus contained a 17-year timeseries of significant wave heights, peak periods and mean wave directions (with both swell and wind waves). This data set is available on a 500m resolution around most of the South African coastline and on a 200m resolution with False Bay as provided by the South African Department of Environmental Affairs (DEA) and Council for Scientific and Industrial Research (CSIR)(@heron2014). The total 17-year average energy, for each output location, was then computed using linear wave theory, taking into account deep, intermediate and shallow water wave approximations (based on wave length) for the wave energy (Holthuijsen, 2007; Kasiulis, Punys and Kofoed, 2015; Foteinis *et al.*, 2017) which is a function of wave height and period.

The directional sheltering effect of the Cape Peninsula, against the dominant swell direction, (given in Figure 1) is clear observed in the wave exposures presented in Figure 2. The classification from fully sheltered to extremely exposed is based on the total wave energy upper and lower limits. It should be noted that what is classified as sheltered around the South African coastline (a high energy coastline) might be classified as exposed in other regions of the world [Norderhaug2012; @Liberti2013; @Sundblad2014]. Kelp and other ecosystems adapt to their typical environmental exposure and thus what is classified as extreme abiotic coastal forcings (temperature, winds, waves and currents) will vary around the world depending on the geographical setting.

@@

In order to compare abiotic parameters for sites around the coast, large historical databases for both temperature and wave energy were accessed. Shallow water temperatures were sourced from the South African Coastal Temperature Network (SACTN) website (https://github.com/ajsmit/SACTN). Seven different organisations within South Africa contribute to the SACTN, where _in situ_ temperature measurements are made around the South African coast using either hand-held thermometers or digital temperature recorders positioned underwater. The mean duration of the 135 daily time series is 19.7 years, and these _in situ_ data are preferred over satellite SST, which have shown to exhibit large biases [@Smit2013]. Linear interpolated SST were calculated for sites where _in situ_ recorders were absent. Wave energy data formed part of the South African Coastal Vulnerability Assessment, presented to the Department of Environmental Affairs [@Theron2014]. These data are first forecasted using NOAA Wave Watch III (WWIII), with National Centers for Environmental Predictions (NCEP) product as the numerical input [@benassai2006]. Hindcast data from WWIII span from 1994-2013 at a 3-hour resolution. The data are then used to model short –crested waves generated by the wind into the coastal environment, using Simulating Waves in the Nearshore [SWAN; @booij1996]. SWAN allows one to extract wave parameters from specific gridded locations in the nearshore. For False Bay, a resolution of 200 meters was modelled, at both 7 meter and 15 meter contours. A 200 meter resolution was used as False Bay was nested within a larger grid area for the research from where the data were sourced.  For Table Bay and east of Cape Hangklip the resolution 500 meters at 7 meters and 15 meters. For this study the 7 meter contours were used.

@@

```
#### Waves as a driver of _Laminaria pallida_ morphometrics

>>>>

Kelps are a group of large seaweeds of the order Laminariales (Ochrophyta), which despite their relatively low taxonomic diversity of species in genera [@Bolton2010], nevertheless form the basis of one of the most productive ecosystems globally [@Mann1973]. Kelps generally have a dependence on cool, temperate and arctic seawater temperatures [@Santelices2007; @Bolton2010], and dominate the nearshore biomass within the rocky shallow coasts in both hemispheres [@Steneck2002]. Outside of these latitudes, they are also found in cool, deep water towards the tropics [@Graham2007]: this deeper tropical water, due to its water clarity, also allows for suitable photon flux and a nitrate concentration [@Luning1990; @Dayton1999; @Zimmerman1984], the two other environmental variables that support kelp productivity. Although kelps have a low taxonomic diversity, their size and complex morphology provide a heterogeneous habitat structure [@Steneck2002] that accommodate a multitude other turf and subcanopy seaweed species, and diverse assemblages of sessile and mobile invertebrates and vertebrates [@Mann1973; @Duggins1989; @Steneck2002], each depending on a wide suite of ecological services provided by the kelp habitat [@Gaines1987; @Bologna1993; @Levin1994; @Anderson1997].

@@

The biogeographic distribution of kelp is limited by seawater temperature [@Bolton2010], where increasing temperature gradients reduce kelp distribution. Due to this limiting factor, the two main species of kelps in southern African waters, _Ecklonia maxima_ and _Laminaria pallida_, are distributed along a section of the south coast from De Hoop, extending west around the Cape Peninsula, and thriving north into Namibia [@Molloy1996; @Stegenga1997]. This distribution follows a temperate gradient, where sea temperatures increase as one moves south from Namibia, around Cape Point and towards De Hoop. Although the two species occur together for the majority of the coast, their basic morphologies and resource needs vary to a degree. The larger species, _Ecklonia maxima_, is distributed from Lüderitz to Cape Agulhas [@Bolton1985a; @Probyn1985; @Bolton1987; @Bolton2012]. Characterised by a large distal swollen bulb filled with gas, and smooth fronds, this species grows to approximately 10 meters [@Bolton1987]. There was, however, a 17-meter specimen collected in 2015 off Cape Point (Smit, unpubl. data).This species of kelp not only dominate the biomass of the South African nearshore, but plays an important ecological role [@Bustamante1996]. The estimated productivity of _Ecklonia maxima_ within South Africa varies between 350 and 1500g Cm-2yr-1 [@Mann1982]. Across the majority of the coastline, _Laminaria pallida_ remains a subsurface kelp, dominating the kelp biomass at depths greater than 10 meters [@Field1980b; @Bolton1987; @Molloy1996]. This species is distributed from Danger Point, east of the Cape Peninsula, to Rocky Point in northern Namibia, and reaches depths of greater than 20 meters [@Field1980b; @Molloy1996; @Molloy1996; @Stegenga1997]. Towards the north along the west coast, from around Hondeklipbaai, _Laminaria pallida_ replaces _Ecklonia maxima_ as the dominant kelp species [@Velimirov1977; @Stegenga1997] and it also occupies increasingly shallower subtidal regions. The northern populations also exhibit an increase in stipe hollowness, compared to the solid stipe morphs in the species’ southern distributions [@Molloy1996]. This variation in morphology was thought to represent two distinct species, with the northern populations formerly described as _Laminaria schinzii_ Foslie [@Molloy1996]. Genetic work has subsequently shown that the two morphs are in fact the same species [@Rothman2017]. In southern African waters, the primary production of _Laminaria pallida_ is between 120 and 1900g C m2yr1, similar to that of _Ecklonia maxima_ [@Mann1982]. Primary production is not the only pathway.

@@

In Figure 2 the total coastal wave exposure of the Cape Peninsula is given in terms of wave energy (kW per meter wave crest length). These values have been computed using model output from approximately 20 year of simulated nearshore wave conditions. Hindcast global wave parameters were refracted into the nearshore via a spectral phase-averaged numerical wave modelling code SWAN (Simulating, WAves in the Nearshore - 3rd generation). The model is fully spectral in all frequencies and directions (0o -360o) (can solve swell and local wind generated waves simultaneously, propagating in different direction) and was used to model the propagation of short crested swell waves, waves generated by wind, non-linear wave-wave interaction and dissipation and depth induced breaking and refraction. Wave energy dissipation due to bottom friction and whitecapping is also accounted for [@Deltares2014].

Time series output from 1997 till 2013 of this model was then produced at the 7m and 15m depth contour lines. Each nearshore output point thus contained a 17-year timeseries of significant wave heights, peak periods and mean wave directions (with both swell and wind waves). This data set is available on a 500m resolution around most of the South African coastline and on a 200m resolution with False Bay as provided by the South African Department of Environmental Affairs (DEA) and Council for Scientific and Industrial Research (CSIR)(@Theron2014). The total 17-year average energy, for each output location, was then computed using linear wave theory, taking into account deep, intermediate and shallow water wave approximations (based on wave length) for the wave energy [@Holthuijsen2007; @Kasiulis2015; @Foteinis2017] which is a function of wave height and period.

The directional sheltering effect of the Cape Peninsula, against the dominant swell direction, (given in Figure 1) is clear observed in the wave exposures presented in Figure 2. The classification from fully sheltered to extremely exposed is based on the total wave energy upper and lower limits. It should be noted that what is classified as sheltered around the South African coastline (a high energy coastline) might be classified as exposed in other regions of the world [@Norderhaug2012; @Liberti2013; @Sundblad2014]. Kelp and other ecosystems adapt to their typical environmental exposure and thus what is classified as extreme abiotic coastal forcings (temperature, winds, waves and currents) will vary around the world depending on the geographical setting.

@@

In order to compare abiotic parameters for sites around the coast, large historical databases for both temperature and wave energy were accessed. Shallow water temperatures were sourced from the South African Coastal Temperature Network (SACTN) website (https://github.com/ajsmit/SACTN). Seven different organisations within South Africa contribute to the SACTN, where _in situ_ temperature measurements are made around the South African coast using either hand-held thermometers or digital temperature recorders positioned underwater. The mean duration of the 135 daily time series is 19.7 years, and these _in situ_ data are preferred over satellite SST, which have shown to exhibit large biases [@Smit2013]. Linear interpolated SST were calculated for sites where _in situ_ recorders were absent. Wave energy data formed part of the South African Coastal Vulnerability Assessment, presented to the Department of Environmental Affairs [@Theron2014]. These data are first forecasted using NOAA Wave Watch III (WWIII), with National Centers for Environmental Predictions (NCEP) product as the numerical input [@benassai2006]. Hindcast data from WWIII span from 1994-2013 at a 3-hour resolution. The data are then used to model short –crested waves generated by the wind into the coastal environment, using Simulating Waves in the Nearshore SWAN [@booij1996]. SWAN allows one to extract wave parameters from specific gridded locations in the nearshore. For False Bay, a resolution of 200 meters was modelled, at both 7 meter and 15 meter contours. A 200 meter resolution was used as False Bay was nested within a larger grid area for the research from where the data were sourced.  For Table Bay and east of Cape Hangklip the resolution 500 meters at 7 meters and 15 meters. For this study the 7 meter contours were used.

@@

#### Waves as a driver of _Laminaria pallida_ morphometrics
